---
title: "Data Science"
author: "Derek Sollberger"
date: "August 29, 2017"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
---

## Welcome to Math 18

Lecturer:  Derek Sollberger

e-mail: dsollberger@ucmerced.edu

office hours: 

* F, 10 AM to 12 noon, SSM 209

## The data

```{r, message = FALSE, warning = FALSE}
library("tidyr")
library("ggplot2")
library("mosaicData")
```

## Graphics in R

1.  base graphics (simplest, but cumbersome)
2.  lattice (usually for chemical studies)
3.  ggplot (used my most data scientists)

## One-Dimensional, Discrete Data

```{r}
data(Dimes)
head(Dimes)
```

## One-Dimensional, Continuous Data

```{r}
data(Dimes)
head(Dimes)
```

## Kernel Density Estimates

```{r}
data(SAT)
head(SAT)
```

## Two, Continuous Variables

```{r}
data(SAT)
head(SAT)
```

## Clicker Question

Were you able to get RStudio running and R Markdown to knit?

1.  Yes
2.  RStudio yes, but not the markdown
3.  No

## Facet Grids

```{r}
data(SwimRecords)
head(SwimRecords)
```

## Color Brewer

```{r}
data(SnowGR)
head(SnowGR)
```

## Example:  Opiate Addiction?

https://www.nytimes.com/interactive/2017/06/05/upshot/opioid-epidemic-drug-overdose-deaths-are-rising-faster-than-ever.html?mcubz=1&_r=0

http://www.worldlifeexpectancy.com/usa-cause-of-death-by-age-and-gender

## Choropleths

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(choroplethrCaCensusTract)
data(df_ca_tract_demographics)
df_ca_tract_demographics$value = df_ca_tract_demographics$median_rent
ca_tract_choropleth(df_ca_tract_demographics,
    title       = "2013 Merced County Census Tracts \n Median Income",
    legend      = "Median Income",
    num_colors  = 5,
    county_zoom = 6047)
```

Source:  https://www.arilamstein.com/blog/2015/09/30/combining-choropleth-maps-and-reference-maps-in-r/

## Choropleths

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(choroplethrCaCensusTract)
data(df_ca_tract_demographics)
df_ca_tract_demographics$value = df_ca_tract_demographics$median_rent
ca_tract_choropleth(df_ca_tract_demographics,
    title       = "2013 Merced County Census Tracts \n Median Income",
    legend      = "Median Income",
    num_colors  = 5,
    county_zoom = 6047,
    reference   = TRUE)
```

## Example:  Cholera Outbreak of 1854

<img src = "cholera.jpg">

## Example:  Cholera Outbreak of 1854

<img src = "cholera.jpg">

Map created by Dr. John Snow

## Looking Ahead

* "Introduction" due Thursday, August 31
* Think about groups

    * who do you want to work with?
    * up to 4 students per group